---
title: "ChlorImport"
output: html_notebook
---

# Introduction
This .rmd imports Turner chlorophyll florescence data from an interactive googlesheet.
It converts RFU from the Turner Fluorometer to ug Chl cm^-2^ of originating tissue paper, used to sample from a model mudflat.
It converts a numeric YYYYMMDDHHMM code to Time of Day, and elapsed time.
It saves the tidied data as a .Rds for general use in the project.


# Install packages if needed
```{r install_packages}
#install.packages("googlesheets4")
#install.packages("googledrive")
#install.packages("tidyverse")
#install.packages("kableExtra")
```


# Loading Packages 
```{r load packages}
library(tidyverse) 
library(googlesheets4)
library(googledrive)
library(kableExtra)
```


# Connect to google account
Deauthorize google account to allow universal access to the raw data sheet.Then retrieve data sheet from where it's stored on google drive.
```{r connect_google}
gs4_deauth()
googlesheet_url <- "https://docs.google.com/spreadsheets/d/16DxWZhYw69ZgACGu82szN9tpgpYYPWIxPXOcSLM2Bo0/edit?gid=1376406540#gid=1376406540"
```

# Accessing sheets
```{r accessing_sheets}
vert_migration <- read_sheet(googlesheet_url, sheet = "vert_migration")

```



# Transforming our numbers into Chl cm^-2^

Instructions for the chlorophyll calculations come from CampbellSOP28ChlorophyllaTurner.pdf
 7.0314 RFU/ug L-1 and intercept of 0.9791
 
```{r chlorophyll_calculations}
vert_migration <- vert_migration |>
  mutate(Chl_ugL = (Fluorescence_Reading/7.0314) -0.9791) |>
  mutate(Chl_ugcm2 = (Chl_ugL*VolSol_mL/1000) / Tissue_Area_cm2) |>
  filter(!is.na(OperatorID)) #get rid of rows that are blanks with OperatorID = NA
  
kable(vert_migration)


#remove garbage columns containing only NA
vert_migration <- vert_migration |>
  select(where(function(x) !all(is.na(x))))

# base R alternate approach from Sarah P.
# chl_data$Chl_ugL <- (chl_data$Fluorescence_Reading/7.0314) -0.9791
# chl_data$Chl_ugcm2 <- (chl_data$Chl_ugL*chl_data$VolSol_mL/1000) / chl_data$Tissue_Area_cm2
# kable(chl_data)
```

# Generate Time of Day & elapsed time
Working with Dates & Times is tricky!
```{r elapsed_time}
vert_migration <- vert_migration |>
  mutate(DateTime = lubridate::ymd_hm(YYYYMMDDHHMM)) |>
  mutate(Date = stringr::str_extract(DateTime, "[0-9]{4}-[0-9]{2}-[0-9]{2}")) |>
  mutate(ToD = stringr::str_extract(DateTime, "[0-9]{2}:[0-9]{2}:[0-9]{2}"))

#hack solution to get elapsed time; difficulty 'subtracting' times
vert_migration <- vert_migration |>
  group_by(Date) |>
  mutate(E_time_h = as.numeric(DateTime - DateTime[1])/3600) |>
  ungroup()


```



```{r save_vert_migration}
saveRDS(object = vert_migration, file = file.path("..", "Data","CleanData", "vert_migration.rds"))
```

