---
title: "ChlorLine"
output: html_notebook
---

# Introduction
This .rmd reads chlorophyll data and generates a line graph ...


# Install packages if needed
```{r install_packages}
#install.packages("tidyverse")
#install.packages("kableExtra")
```


# Loading Packages 
```{r load packages}
library(tidyverse) 
library(kableExtra)
```


```{r read_vert_migration}
vert_migration <- readRDS( file = file.path("..","Data","CleanData","vert_migration.rds"))
```

```{r chlor_time}
vert_migration |>
  ggplot() +
  geom_point(aes(x=ToD, y = Chl_ugcm2 )) + 
  facet_grid(rows = vars(Top_Bottom), cols = vars(Temp_C))
```
# Create top and bottom ratio
vert_migration contains variable Top_Bottom
To simplify ratio calculation we 'pivot_wider' the data to separate Top_Bottom into two separate variables, Top & Bottom; that will lower the number of rows and increase the number of columns
```{r top_bottom}
vert_migration_wide<- vert_migration |>
  pivot_wider(vert_migration, id_cols = names_from = Top_Bottom, values_from = Chl_ugcm2)


```



