---
title: "ChlorImport"
output: html_notebook
---

# Introduction
This .rmd imports chlorophyll florescence data ...


# Install packages if needed
```{r install_packages}
#install.packages("googlesheets4")
#install.packages("googledrive")
#install.packages("tidyverse")
#install.packages("here")
#install.packages("kableExtra")
```


# Loading Packages 
```{r load packages}
library(tidyverse) 
library(here)
library(googlesheets4)
library(googledrive)
library(kableExtra)
```


# Connect to google account
Deauthorize google account to allow universal access to the raw data sheet.Then retrieve data sheet from where it's stored on google drive.
```{r connect_google}
gs4_deauth()
googlesheet_url <- "https://docs.google.com/spreadsheets/d/16DxWZhYw69ZgACGu82szN9tpgpYYPWIxPXOcSLM2Bo0/edit?gid=1376406540#gid=1376406540"
```

# Accessing sheets
```{r accessing_sheets}
vert_migration <- read_sheet(googlesheet_url, sheet = "vert_migration")
```



# Transforming our numbers into Chl/cm^2
We need to make it so that we are looking directly at how much chlorophyll per cm^2. I don't know how to do this calculation with VolSol.
I am going to make preliminary graphs by dividing off the Tissue Area and timesing by VolSol. When we know correct calculations, we can change this.
This will go into a new column called Chl_cm2
Instructions for the chlorophyll calculations come from CampbellSOP28ChlorophyllaTurner.pdf
 7.0314 RFU/ug L-1 and intercept of 0.9791
 
```{r chlorophyll_calculations}
vert_migration <- vert_migration |>
  mutate(Chl_ugL = (Fluorescence_Reading/7.0314) -0.9791) |>
  mutate(Chl_ugcm2 = (Chl_ugL*VolSol_mL/1000) / Tissue_Area_cm2) 
  
kable(vert_migration)

# chl_data$Chl_ugL <- (chl_data$Fluorescence_Reading/7.0314) -0.9791
# chl_data$Chl_ugcm2 <- (chl_data$Chl_ugL*chl_data$VolSol_mL/1000) / chl_data$Tissue_Area_cm2
# kable(chl_data)
```

```{r save_vert_migration}
saveRDS(object = vert_migration, file = file.path("..","Code","CleanData","vert_migration.rds"))
```

