---
title: "R Notebook"
output: html_notebook
---


# Loading Packages 
```{r}
#install.packages("tidyverse", "knitr", "here", "ggplot2")
library(tidyverse) 
library(knitr)
library(here)
library(ggplot2)
#library(dplyr)
```

# Data Dictionary
Loading in the data from our personal computer files. Edit here if we need to switch what data is being used.
```{r}
DataDictionary <-read_csv(file = here("Data", "RawData", "vert_migration.csv"))
kable(DataDictionary)
```
#Transforming our numbers into Chl/cm^2
We need to make it so that we are looking directly at how much chlorophyll per cm^2. I don't know how to do this calculation with VolSol.
I am going to make preliminary graphs by dividing off the Tissue Area and timesing by VolSol. When we know correct calculations, we can change this.
This will go into a new column called Chl_cm2
```{r}
DataDictionary$Chl_cm2 <- (DataDictionary$Fluorescence_Reading*DataDictionary$VolSol_mL) / DataDictionary$Tissue_Area_cm2
kable(DataDictionary)
```

#Sorting the data, creating lists 
To make the data usable for future. I need the categories which will become each bar on a bar plot. For my first plot, I am going with top sheets comparing all conditions.

The for loops and lists has it so that it will cycle through each of the different conditions. When each condition is present, I can manipulate it to a list as the std and the mean. I will need the std and the mean for the bar graph. 
```{r}
temp <- c(20, 15)
sheet <- "top"
condition <-c("dark", "med", "light")
time <- 202410101430

top_means <-c()
top_std <-c()
top_conditions<-c()

for (y in 1:2) {
  for (x in 1:3) {
    sub_data <- filter(DataDictionary, Temp_C == temp[[y]], Top_Bottom == sheet, Light == condition[[x]], Chl_cm2 > 0, YYYYMMDDHHMM == time)
    top_means <-append(top_means, mean(sub_data$Chl_cm2))
    top_std <-append(top_std, sd(sub_data$Chl_cm2))
    top_conditions <-append(top_conditions, paste(temp[[y]], condition[[x]]))
  }
}

kable(sub_data)
#print(top_std)
```

#Bar Graph 
```{r}
data <- data.frame(top_conditions, top_means)
ggplot(data, aes(x=top_conditions, y=top_means)) + 
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin=top_means-top_std, ymax=top_means+top_std), width=.2,
                 position=position_dodge(.9))
 
```

