---
title: "Mud Migration Community Compositon"
output: html_notebook
author:
  Paige Stuart
  Douglass Campbell
---
# Install packages if needed
```{r install_packages}
#install.packages("googlesheets4")
#install.packages("googledrive")
#install.packages("tidyverse")
```

# Load packages
```{r load_packages}
library(googlesheets4)
library(googledrive)
library(tidyverse)
```

# Connect to google account
De-authorize Google account to allow universal access to the raw data sheet.Then retrieve data sheet from where it's stored on google drive.
```{r connect_google}
gs4_deauth()
googlesheet_url <- "https://docs.google.com/spreadsheets/d/1Ga0qnVxOVR9NwSNI5RM1iz1L2MoCAyF0O4x3D6mCePw/edit?gid=405129858#gid=405129858"
```

# Accessing sheets
```{r accessing_sheets}
#Accessing sheet with meta data
community_composition <- read_sheet(googlesheet_url, sheet = "community_composition") 
#Accessing sheet with raw data 
raw_cell_measurements <- read_sheet(googlesheet_url, sheet = "raw_cell_measurements") 
#Accessing data dictionary for reference to column values
data_dictionary <- read_sheet(googlesheet_url, sheet = "data_dictionary")
```
# Set Up
Initial steps for setting up the raw data for analysis. 

## Set missing values of width_1 and width_2
Some width measurements are blank is a) the cell was undefined or b) the morphology had no unequal widths.
```{r missing_widths}
#Correcting width 1
corrected_width_data <- raw_cell_measurements |>
  mutate(Width1_um = case_when(
    is.na(Width1_um) ~ 1,
    TRUE ~ Width1_um
  ))
#Correcting width 2
corrected_width_data <- corrected_width_data |>
  mutate(Width2_um = case_when(is.na(Width2_um) ~ Width1_um,
                               !is.na(Width2_um) ~ Width2_um
  ))
```

## List of morphotypes obseved 
Before running any code on morphotypes I wanted to check for any misspelled words.
```{r unqiue_morphotypes}
unique_morphotypes <- unique(corrected_width_data$CellMorphotype)
print(unique_morphotypes)
```

## Correct mispelled pennate
A single morphotype label was misspelled so I corrected it.
```{r correct_mispelling}
#Find cell containing the misspelled word
mislabel_pennate <- which(corrected_width_data  == "penate", arr.ind = TRUE)
mislabel_pennate

#Indicate the cell which was mislabeled
corrected_width_data[106, "CellMorphotype"] <- "pennate"
```

## Fix missing CellVol
One cell did not fit well into any of the volume calculations 
```{r fix_missing}
#Rename morphotype to better calculate cell volume
corrected_width_data[12, "CellMorphotype"] <- "other"
```

## Estimate volumes & maximum linear dimension
NEED DISCRIPTION - DELETE VOLUME?
```{r volumes_dimension}
#Calculate Cell Volume for each Morphotype
volume_dimension_data <- corrected_width_data |>
  mutate(CellVol = case_when(CellMorphotype == "centric" ~ pi * (Width1_um/2)^2 *  Width2_um,
                          CellMorphotype == "pennate" ~  Length_um * Width1_um * Width2_um,
                          CellMorphotype == "undefined" ~ 4/3 * pi * (Length_um/2)^3, 
                          CellMorphotype == "peanut" ~   Length_um * Width1_um * Width2_um,
                          CellMorphotype == "rectangular" ~ Length_um * Width1_um * Width2_um,
                          CellMorphotype == "square"~ Width1_um^2 * Width2_um, #check these values
                          CellMorphotype == "other" ~ 4/3 * pi * (Width1_um/2)^3
  )
  )

#Calculate Maximum Dimension
volume_dimension_data <- volume_dimension_data |>
  rowwise() |>
  mutate(MaxDim_um = max(Length_um, Width1_um, Width2_um, na.rm = TRUE)
  )
```

## Check for missing results
NEED DISCRIPTION - DELETE VOLUME?
```{r missing_results}
#Find empty cells in CellVol
#empty_CellVol <- which(is.na(volume_dimension_data$CellVol))
#volume_dimension_data[empty_CellVol, ]

#Find it's location
#empty_CellVol_location <- cbind(row = empty_CellVol, column = "CellVol")
#empty_CellVol_location

#Find empty cells in MaxDim_um
#empty_MaxDim <- which(is.na(volume_dimension_data$MaxDim_um))
#volume_dimension_data[empty_MaxDim, ]

#Corrected these missing results, return to misspelling above ^
```

## Control for chain sizes
NEED DISCRIPTION - DELETE VOLUME?
This conversion works because, chains all centric
```{r size_chain}
volume_dimension_data <- volume_dimension_data %>%
  mutate(CellVol_Total = CellVol * ChainNum) %>%
  mutate(MaxDim_Total = MaxDim_um * ChainNum)
```

## Histogram of max dimension
```{r histogram}
#Plot Maximum Dimensions
volume_dimension_data |>
  ggplot()+
  geom_histogram(aes(MaxDim_Total))
```

## Merge common value between raw data and conditions (light, temp, layer)
```{r merge_sheets}
merged_data <- merge(community_composition, volume_dimension_data, by = "SamplePhotoID", all.x = TRUE)
```

```{r plot max dim}
merged_data |>
  ggplot()+
  geom_histogram(aes(MaxDim_Total)) +
    labs(
    x = "Maximum Dimension (um³)", 
    y = "Cell Count per 0.32 mm2", 
    title = "Cell Count per Condition and Maximum Dimension"
  ) +
  facet_grid(rows = vars(Top_Bottom), cols = vars(Temp_C, Light)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), 
    axis.title.x = element_text(size = 12))
```

# Cell Count Analysis

## Create bargraph to visualize cell counts per sample
```{r bargraph_sample}
#Summarize data with chain number
merged_data_cell_count <- merged_data %>%
  group_by(Light, Temp_C, Top_Bottom, SamplePhotoID) %>%
  summarize(CellCount = sum(ChainNum), .groups = "drop")

#Create bar graph
ggplot(merged_data_cell_count, aes(x = SamplePhotoID, y = CellCount)) +
  geom_bar(stat = "identity", fill = "purple") +
  labs(x = "Sample", y = "Number of Cells per 0.16mm2", title = "Number of Cells per Sample") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
### Add caption for each figure to describe the unit on the y-axis, for future review 
#### Can reorder for bottom on the bottom and dark, med, light

## Count number of observation in each maximum dimension bin for ONLY pennate and centric diatoms per condition
```{r filtered_data}
 #Summarize the counts for each maximum dimension bin per condition
filtered_data <- merged_data |>
  dplyr::filter(CellMorphotype %in% c("pennate", "centric")) |>
  dplyr::group_by(Patch, Light, Temp_C, Top_Bottom, CellMorphotype, ChainNum, MaxDim_Total)

#print(filtered_data)
```

## Reorder top and bottom
```{r reorder_top_bottom}
filtered_data$Top_Bottom <- factor(filtered_data$Top_Bottom, levels = c("top", "bottom"))
```

# Cell count per maximum dimension and condition
```{r count_dimension}
filtered_data |>
  ggplot() +
  geom_histogram(aes(x = MaxDim_Total, fill = Light), bins = 30, color = "black", position = "stack") +
  labs(
    x = "Maximum Dimension (um³)", 
    y = "Cell Count per 0.32 mm²", 
    title = "Total Cell Count per Condition and Maximum Dimension"
  ) +
  facet_grid(rows = vars(Top_Bottom), cols = vars(Temp_C)) + 
  scale_fill_brewer(palette = "Set1") + 
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), 
    axis.title.x = element_text(size = 12),
    strip.text.y = element_text(size = 12), 
    strip.text.x = element_text(size = 14), 
    legend.position = "bottom", 
    strip.text = element_blank() 
  )
```

# Cell count per condition 
```{r cell_count}
cell_count_data <- filtered_data %>%
  group_by(Light, Temp_C, Top_Bottom, SamplePhotoID) %>%
  summarize(CellCount = sum(ChainNum), .groups = "drop")

ggplot(cell_count_data, aes(x = Light, y = CellCount, fill = Light)) +
  geom_bar(stat = "identity") +
  labs(x = "Patch (Corresponding Top and Bottom Samples)", 
       y = "Cell Count per Image (0.32 mm3)", 
       title = "Cell Count per Condition") +
  facet_grid(rows = vars(Top_Bottom), cols = vars(Temp_C)) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),  
        axis.title.x = element_text(size = 12))  
```

# Top-bottom ratio graph per patch and condition
ASK DR.C ABOUT RATIO CALCULATION
```{r top_bottom_ratio}
# Ratio of Top and Bottom counts within each Patch
top_bottom_ratio_data <- filtered_data |>
  dplyr::filter(Top_Bottom %in% c("top", "bottom")) |> 
  dplyr::group_by(Patch, Top_Bottom, Light, Temp_C) |> 
  dplyr::summarize(Total_Count = sum(ChainNum), .groups = "drop") |> 
  tidyr::pivot_wider(names_from = Top_Bottom, values_from = Total_Count, values_fill = 0) |>
  dplyr::mutate(Ratio = top / bottom)
  #dplyr::mutate(Ratio = bottom / top)

ggplot(top_bottom_ratio_data, aes(x = Patch, y = Ratio, fill = Light)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    x = "Patch (Corresponding Top and Bottom Samples)", 
    y = "Top-to-Bottom Ratio of Cell Counts per 0.32 mm2",
    title = "Top-to-Bottom Ratio with Temperature and Light Conditions"
  ) +
  facet_wrap(~ Temp_C) + 
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


















