---
title: "Mud Migration Community Compositon"
output: html_notebook
---
# Install packages if needed
```{r install_packages}
#install.packages("googlesheets4")
#install.packages("googledrive")
#install.packages("tidyverse")
```

# Load packages
```{r load_packages}
library(googlesheets4)
library(googledrive)
library(tidyverse)
```

# Connect to google account
Deauthorize google account to allow universal access to the raw data sheet.Then retrive data sheet from where it's stored on google drive.
```{r connect_google}
gs4_deauth()
googlesheet_url <- "https://docs.google.com/spreadsheets/d/1Ga0qnVxOVR9NwSNI5RM1iz1L2MoCAyF0O4x3D6mCePw/edit?gid=405129858#gid=405129858"
```

# Accessing sheets
```{r accessing_sheets}
community_composition <- read_sheet(googlesheet_url, sheet = "community_composition")
raw_cell_measurements <- read_sheet(googlesheet_url, sheet = "raw_cell_measurements")
data_dictionary <- read_sheet(googlesheet_url, sheet = "data_dictionary")
```

# Set missing width_1 values to equal 1um 
```{r mutate_width1}
raw_cell_measurements_2 <- raw_cell_measurements |>
  mutate(Width1_um = case_when(
    is.na(Width1_um) ~ 1,
    TRUE ~ Width1_um
  ))

raw_cell_measurements_3 <- raw_cell_measurements_2 |>
  mutate(Width2_um = case_when(is.na(Width2_um) ~ Width1_um,
                               !is.na(Width2_um) ~ Width2_um
  ))

```

# Estimate Volumes & Maximum Linear Dimension

```{r volumes_maxlinear}

#need volume equations for cylinder (Centric), rectangular prism (pennate), oblate spheroid for undefined & peanut?

raw_cell_measurements_4 <- raw_cell_measurements_3 |>
  mutate(CellVol = case_when(CellMorphotype == "centric" ~ pi * (Width1_um/2)^2 *  Width2_um,
                          CellMorphotype == "pennate" ~  Length_um * Width1_um * Width2_um,
                          CellMorphotype == "undefined" ~ 4/3 * pi * (Length_um/2)^3, 
                          CellMorphotype == "peanut" ~   Length_um * Width1_um * Width2_um
  )
  )

raw_cell_measurements_4 <- raw_cell_measurements_4 |>
  rowwise() |>
  mutate(MaxDim_um = max(Length_um, Width1_um, Width2_um, na.rm = TRUE)
  )
```

# Histogram of volumes and max dimension
```{r histogram}
raw_cell_measurements_4 |>
  ggplot()+
  geom_histogram(aes(CellVol))

raw_cell_measurements_4 |>
  ggplot()+
  geom_histogram(aes(MaxDim_um))
```


#Bin surface areas
```{r bin_areas}
# equal breaks
#raw_cell_measurements_2$Surface_Area_Bins <- cut(raw_cell_measurements_2$Surface_Area, breaks = 5, labels = FALSE)

# custom bin ranges
#bin_ranges <- c(0, 10, 20, 30, 40, 50)
#raw_cell_measurements_2$Surface_Area_Bins <- cut(raw_cell_measurements_2$Surface_Area, breaks = bin_ranges, include.lowest = TRUE)
```

#Merge common value between raw data and conditions
```{r merge_sheets}
#merged_data <- merge(community_composition, raw_cell_measurements_4, by = "SamplePhotoID", all.x = TRUE)

#merged_data_2 <- left_join(raw_cell_measurements_4, community_composition, by = "SamplePhotoID")

merged_data_3 <- right_join(community_composition, raw_cell_measurements_4, by = "SamplePhotoID")
```

